---
title: Measuring NC job loss from Covid-19
author: Frank Muraca
date: '2020-04-25'
slug: measuring-nc-job-loss-from-covid-19
categories: []
tags: []
Description: ''
Tags:
  - Covid-19
  - Economic Development
---
```{r, echo=FALSE}
knitr::knit_hooks$set(
   message = function(x, options) {
     paste('\n\n<div class="alert alert-info">',
           gsub('##', '\n', x),
           '</div>', sep = '\n')
   }
)


```

Over the past four weeks, more than 20 million people [have filed for unenmployment](https://www.nytimes.com/2020/04/16/business/economy/unemployment-numbers-coronavirus.html) as the country adjusts to the economic fallout of Covid-19. While we still don't know the longterm impacts of the pandemic, widespread job loss creates an immediate need for Americans without savings to cover their cost of living. 

What do we know about the people and places that have been most impacted by the recent shock in unemployment?

In mid-April the Urban Institute [released new estimates](https://datacatalog.urban.org/dataset/estimated-low-income-jobs-lost-covid-19) on the number of low-income jobs (<$40k salary) lost as a result of Covid-19. In this post, we'll use this data to figure out where job loss is most prevalent in North Carolina. 

## The data

The estimates rely on unemployment reports from Washington State and New York State that include job losses by industry. To calculate job loss for Census tracts elsewhere in the country, those estimates are applied to state-level job loss estimates reported by the Bureau of Labor Statistics. The full methodology is [here.](https://www.urban.org/sites/default/files/where_low-income_jobs_are_being_lost_to_covid-19_technical_appendix_04232020.pdf)

The Urban Institute provides two separate files for county and tract estimates. Let's start with counties:

```{r, message=FALSE, warning=FALSE}
# We'll be using the following packages:
library(geojsonsf)
library(sf)
library(tidyverse)
library(hrbrthemes)
library(scales)
```

```{r, message=FALSE, warning=FALSE}
# Let's pull in the county-level data and subset the file to only include counties in North Carolina. We'll also remove the word "County" from the names to make it more readable later.

county <- geojson_sf("https://ui-lodes-job-change-public.s3.amazonaws.com/sum_job_loss_county.geojson") %>% 
  filter(state_name=="North Carolina") %>%
  mutate(county_name = str_remove(county_name," County"))

colnames(county)

# The industries are coded as "X01", "X02", "X03", and so on. The Urban Institute also provides a codebook that we'll use to rename each of the 21 industries. 

industry_codebook <- read_csv("https://ui-lodes-job-change-public.s3.amazonaws.com/lehd_types_s3.csv")

## Break out industry variables into new dataframe
industry <- county %>% select(starts_with("X"), -geometry)  

## Replace industry variable names based on matched name from the codebook
names(industry) <- industry_codebook$lehd_name[match(names(industry),industry_codebook$lehd_var)]

county <- county %>% 
  select(-starts_with("X")) %>% ## drop the uncoded variables
  cbind(industry) ## merge columns with the new renamed variables

library(janitor)
county <- county %>% clean_names()

```

## Where are North Carolina's 'hadest hit' counties?

Let's visualize the top 15 counties by total job loss using a [lollipop plot.](https://www.r-graph-gallery.com/lollipop-plot.html) 

```{r, warning=FALSE, message=FALSE}
county_viz_losstotal <- county %>%
  arrange(desc(total_job_loss_index)) %>% ## put highest job loss counties at the top
  top_n(total_job_loss_index,n=15) %>% ## select top 15
  select(county_name,total_job_loss_index) %>%
  ggplot(aes(x=reorder(county_name,total_job_loss_index),y=total_job_loss_index)) + # have our counties descend by total job loss count
  geom_point() +
  geom_segment(aes(x=county_name,xend=county_name,y=0,yend=total_job_loss_index)) +
  coord_flip() +
  theme_minimal() +
  scale_y_comma(limits=c(0,52000)) +
  labs(y="Total Job Loss",
       title="Estimated Job Loss from Covid-19",
       subtitle = "Top 15 Counties in NC, Jobs w/ salaries <$40k",
       caption="@FrankMuraca, Data Source: Urban Institute",
       x=NULL) 
county_viz_losstotal
```

Unsurprisingly, job loss is highest in North Carolina's seven metro counties - with Raleigh and Charlotte accounting for about **20%** of jobs lost in the state. Now let's see if the same is true when looking at job loss as a **percentage of the overall labor force.** 

Let's use the [`tidycensus`](https://walkerke.github.io/tidycensus/articles/basic-usage.html) package to pull labor force estimates for each county in the state. If you're a longtime American Factfinder user ([RIP](https://data.census.gov/cedsci/deeplinks?url=https%3A%2F%2Ffactfinder.census.gov%2F&tid=GOVSTIMESERIES.CG00ORG01)), this is a great tool to quickly grab Census data over many years and for different geographies.

```{r, echo=FALSE}
message("**Note**: The Tidycensus requires an API key, [which can be acquired here.](https://api.census.gov/data/key_signup.html)")

```

```{r, message=FALSE,warning=FALSE,echo=FALSE}
library(tidycensus)
census_api_key("ad2c6554a53beeff954ff8a4b291b02f053d3d31",overwrite = TRUE)

```

```{r, message=FALSE, warning=FALSE}
library(tidycensus) 

county_laborforce <- get_acs(variables = c(civilian_lf = "B23025_003"), ## The variable code for civilian labor force
                             year = 2018,
                             state="NC",
                             geography="county") %>%
  mutate(NAME=str_remove(NAME, " County, North Carolina")) ## Clean up county name column so we can merge with the job loss dataframe

county <- merge(county, county_laborforce, by.x='county_name', by.y ='NAME') ## Merge with original county data

## Recreate job loss by county but as percentage of overall labor force
county_viz_losslbf <- county %>%
  mutate(pctlbf = total_job_loss_index/estimate) %>% ## "Estimate" stands in for labor force when pulled using TidyCensus
  arrange(desc(pctlbf)) %>%
  top_n(pctlbf,n=15) %>%
  ggplot(aes(x=reorder(county_name,pctlbf),y=pctlbf)) +
  geom_point() +
  geom_segment(aes(x=county_name,xend=county_name,y=0,yend=pctlbf)) +
  coord_flip() +
  theme_minimal() +
  scale_y_percent(limits = c(0,.20)) +
  labs(y="Job Loss as Percentage of Labor Force",
      title="Estimated Job Loss from Covid-19 (% of Labor Force)",
       subtitle = "Top 15 Counties in NC, Jobs w/ salaries <$40k",
       caption = "@FrankMuraca, Data Source: Urban Institute",
       x=NULL) 

county_viz_losslbf

```

When we look at job loss as percentage of overall labor force, North Carolina's more rural counties are hardest hit by the economic fall out of Covid-19. Let's use the `tmap` package to visualize the difference in these two metrics.

```{r,message=FALSE,warning=FALSE}
library(tmap)

## Create map of total loss
map_county_totalloss <- county %>%
  tm_shape() +
  tm_polygons(c("total_job_loss_index"), # Variable we want to visualize
              title = c("Total Job Loss by County"),
              style= "cont",
              palette = "seq",
              border.col = NULL) + 
  tm_legend(legend.position = c("left", "top"))

# Create map of job loss as percentage of labor force
map_county_pctlbf <- county %>%
  mutate(pctlbf = (total_job_loss_index/estimate)*100) %>%
  tm_shape() +
  tm_polygons(c("pctlbf"), # Variable we want to visualize
              title = c("Job Loss as % of Labor Force"),
              style= "cont",
              palette = "seq",
              border.col = NULL,
              legend.format=c(list(fun=function(x) paste0(formatC(x, digits=0, format="f"), "%")))) + # To convert numbers to percentages in the legend
  tm_legend(legend.position = c("left", "top"))

library(grid) # Use the grid package the stack the two maps
grid.newpage()
pushViewport(viewport(layout = grid.layout(2,1)))
print(map_county_totalloss,vp=viewport(layout.pos.row = 1))
print(map_county_pctlbf,vp=viewport(layout.pos.row = 2))


```

## What neighborhoods are hardest hit?

Not all industries are equally impacted by the pandemic. Food, hospitality, retail, and some health care subsectors have had to cut back or stop operations to help slow the spread of Covid-19. In North Carolina, over 140,000 of the low-income jobs lost occurred in food & accomodation industries, and nearly 90,000 were lost in retail industries.

```{r,echo=TRUE,warning=FALSE,message=FALSE}
county %>%
  select(county_name, county_fips, state_name, variable, estimate, moe, max, tmax, everything()) %>% # We'll rearrange the columns to keep industries together
  pivot_longer(cols=accomodation_and_food_services:public_administration, # Put industry columns into "long" format
               names_to = "Industry", # Create new column that will contain industry names
               values_to = "Job_Loss") %>% # New column for job loss estimates
  # Let's clean up the industry names 
  mutate(Industry = str_replace_all(Industry,"_"," "), # replace underscores with spaces
         Industry = str_replace_all(Industry, "and", "&"), # replace 'and" with '&'
         Industry = str_to_title(Industry), # capitalize industry names
         Industry = recode(Industry, # Shorten admin support services industry name
                           'Administrative & Support & Waste Management & Remediation Services' = "Admin & Support Services")) %>%
  group_by(Industry) %>%
  filter(Industry != "Total Job Loss Index") %>% # remove total jobs to avoid double counting
  summarise(totaljobs = sum(Job_Loss)) %>%
  ggplot(aes(x=reorder(Industry,totaljobs),y=totaljobs)) +
  geom_point() +
  geom_segment(aes(x=Industry,xend=Industry,y=0,yend=totaljobs)) +
  coord_flip() +
  theme_minimal() +
  scale_y_comma(limits = c(0,150000),
                breaks = seq(0,140000,20000)) +
  labs(x="",
       y="Total Job Loss",
       title="Estimated Job Loss from Covid-19: North Carolina",
       subtitle = "By Industry, Jobs w/ salaries <$40k",
       caption = "@FrankMuraca, Data Source: Urban Institute")


```

Up until now, we've been looking at state and county-wide job loss. What about at the neighborhood level? The Urban Institute also provides estimates for low-income job loss for Census tracts. Let's download that data and look the relationship between job loss and tract poverty rate. Many of these steps will be repeated from the county data. 

```{r setup & cleaning, warning=FALSE, message=FALSE}
tract <- geojson_sf("https://ui-lodes-job-change-public.s3.amazonaws.com/job_loss_by_tract.geojson") %>% filter(substr(GEOID,1,2)==37)

### Filter and rename variables based on codebook --------
industry <- tract %>% select(starts_with("X"), -geometry) ## Break out industry variables into new dataframe

names(industry) <- industry_codebook$lehd_name[match(names(industry),industry_codebook$lehd_var)] ### Replace industry variable names based on matched name from the codebook

tract <- tract %>% 
  select(-starts_with("X")) %>% ## drop the uncoded variables
  cbind(industry) ## merge columns with the new renamed variables

tract <- tract %>% clean_names()
```

```{r merge with census, warning=FALSE,message=FALSE}
### Select census data for NC census tracts
tract_pov <- get_acs(variables = c(population = "B17010_001", # total population
                                   belowpov = "B17010_002", # population below poverty rate
                                   civilian_lf = "B23025_003"), #civilian labor force
                             year = 2018,
                             state="NC",
                             geography="tract") %>%
  pivot_wider(id_cols=GEOID,  ## Pivot data frame to "wide" format so we can merge with the job loss dataframe
              names_from = variable,
              values_from = estimate) %>%
  mutate(povrate = belowpov/population)

tract <- merge(tract,tract_pov,
                        by.x = "geoid",by.y="GEOID") %>%
  mutate(pctlbf = total_job_loss_index/civilian_lf,
         food_pctlbf = accomodation_and_food_services/civilian_lf)

tract %>%
  filter(pctlbf<.3, # remove outliers
         povrate>0) %>%
  ggplot(aes(x=povrate,y=pctlbf)) + 
  geom_point(alpha=.3) +
  geom_smooth(method = "loess") +
  theme_minimal() +
  scale_x_percent(breaks=seq(0,.7,.1)) +
  scale_y_percent() +
  labs(x="Tract Poverty Rate",
       y="Job Loss as Percentage of Tract Labor Force",
       title = "Estimated Job Loss from Covid-19: North Carolina Census Tracts",
       subtitle="Total jobs w/ salaries <$40k as percentage of labor force",
       caption = "@FrankMuraca, Data Source: Urban Institute")

```

Census tracts with higher poverty rates tend to have higher levels of low-income job loss. Based on research from other counties, we know that low-income individuals [are more likely to contract and die from Covid-19](https://www.nytimes.com/2020/03/15/world/europe/coronavirus-inequality.html). This initial look at job loss also shows that the economic impacts of the virus are also more severe in high-poverty areas. This insight should help policymakers direct both *health* and *economic recovery* resources [to these hardest hit communities.](https://shelterforce.org/2020/03/24/housing-policy-must-change-in-wake-of-covid-19/)  